---
title: "Money and Success in Football"
author: "Md. Muktadir"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    toc: true
    toc_depth: '3'
    latex_engine: xelatex
    number_sections: true
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
    fig_width: 10
    fig_height: 6
    df_print: paged
    number_sections: true
subtitle: |
  Does Financial Strength Predict League Position?
  *Sports Economics Seminar — Summer Semester 2026*
---

```{css, echo=FALSE}
/* ── Project colour palette ───────────────────────────────── */
:root {
  --navy:   #1A2A4A;
  --gold:   #F5A623;
  --red:    #C0392B;
  --green:  #196F3D;
  --steel:  #2C3E73;
  --slate:  #566573;
}

/* ── Callout boxes ────────────────────────────────────────── */
.callout-finding {
  background: #FEF9E7;
  border-left: 5px solid #F5A623;
  padding: 14px 18px;
  margin: 16px 0;
  border-radius: 0 6px 6px 0;
  font-size: 0.96em;
}
.callout-alert {
  background: #FDEDEC;
  border-left: 5px solid #C0392B;
  padding: 14px 18px;
  margin: 16px 0;
  border-radius: 0 6px 6px 0;
  font-size: 0.96em;
}
.callout-stat {
  background: #EBF5FB;
  border-left: 5px solid #2C3E73;
  padding: 14px 18px;
  margin: 16px 0;
  border-radius: 0 6px 6px 0;
  font-size: 0.96em;
}
.callout-success {
  background: #EAFAF1;
  border-left: 5px solid #196F3D;
  padding: 14px 18px;
  margin: 16px 0;
  border-radius: 0 6px 6px 0;
  font-size: 0.96em;
}

/* ── Table styling ────────────────────────────────────────── */
table.table thead tr th {
  background-color: #1A2A4A;
  color: white;
}
.best-row { font-weight: bold; background: #FEF9E7 !important; }

/* ── Misc ─────────────────────────────────────────────────── */
h1, h2, h3 { color: #1A2A4A; }
.caption    { color: #566573; font-style: italic; font-size: 0.88em; }
```

```{r knitr-setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  message   = FALSE,
  warning   = FALSE,
  fig.align = "center",
  fig.width = if (knitr::is_latex_output()) 7   else 10,
  fig.height= if (knitr::is_latex_output()) 4.5 else 6,
  dpi       = 150,
  comment   = "#>"
)
options(knitr.kable.NA = "", scipen = 6, digits = 4)
```

```{r global-load, include=FALSE}
# ── Source project utilities ──────────────────────────────────
source("utils/packages.R")
source("utils/constants.R")
source("utils/helpers.R")
load_packages(c("dplyr", "readr", "ggplot2", "broom",
                "tidyr", "scales", "purrr", "knitr", "forcats"))

ensure_installed <- function(pkgs) {
  miss <- pkgs[!vapply(pkgs, requireNamespace, logical(1), quietly = TRUE)]
  if (length(miss))
    install.packages(miss, repos = "https://cloud.r-project.org", quiet = TRUE)
  invisible(lapply(pkgs, library, character.only = TRUE))
}
ensure_installed(c("kableExtra", "ggrepel"))

# kableExtra wrapper: HTML gets Bootstrap styling; PDF avoids tabu bug via full_width=TRUE
kbs <- function(tbl, fw = FALSE, pos = "center",
                bo = c("striped","hover","condensed","responsive"), ...) {
  if (knitr::is_latex_output()) {
    kable_styling(tbl, full_width = FALSE,
                  latex_options = c("repeat_header", "striped"), ...)
  } else {
    kable_styling(tbl, bootstrap_options = bo,
                  full_width = fw, position = pos, ...)
  }
}

# ── Project colour palette ────────────────────────────────────
league_colours <- c(
  "Bundesliga"     = "#E41A1C",
  "La Liga"        = "#FF7F00",
  "Ligue 1"        = "#4DAF4A",
  "Premier League" = "#984EA3",
  "Serie A"        = "#377EB8"
)
league_order <- names(league_colours)

# ── Load master dataset ───────────────────────────────────────
df <- read_csv("data/merged/master_dataset.csv", show_col_types = FALSE) |>
  filter(!is.na(League_Position), !is.na(Financial_Rank),
         !is.na(Market_Value_M_EUR), !is.na(Log_Normalized_Value)) |>
  mutate(
    League      = factor(as.character(League), levels = league_order),
    Champion_bin = as.integer(
      tolower(as.character(Champion)) %in% c("yes", "true", "1")
    )
  )

# ── Fit key models ────────────────────────────────────────────
modA <- lm(League_Position ~ Financial_Rank, data = df)
modB <- lm(League_Position ~ Financial_Rank + League, data = df)
modC <- lm(League_Position ~ Log_Normalized_Value, data = df)
modD <- lm(League_Position ~ Log_Normalized_Value + League + Season_Index, data = df)
modH <- lm(League_Position ~ Financial_Rank + I(Financial_Rank^2), data = df)
modE <- glm(Champion_bin ~ Financial_Rank, data = df, family = binomial())

# ── Augment Model H ───────────────────────────────────────────
aug_H <- augment(modH, data = df) |>
  rename(Fitted_H = .fitted, Residual_H = .resid,
         CooksD = .cooksd, Hat = .hat)

resid_sd <- sd(aug_H$Residual_H, na.rm = TRUE)

aug_H <- aug_H |>
  mutate(
    Outlier_Flag = case_when(
      Residual_H < -1.5 * resid_sd ~ "Overperformer",
      Residual_H >  1.5 * resid_sd ~ "Underperformer",
      TRUE                          ~ "Normal"
    )
  )

# ── Pre-compute inline stats ──────────────────────────────────
n_obs      <- nrow(df)
r_pearson  <- round(cor(df$Financial_Rank, df$League_Position,
                        use = "complete.obs"), 3)
r2_H       <- round(summary(modH)$r.squared, 3)
adj_r2_H   <- round(summary(modH)$adj.r.squared, 3)
aic_H      <- round(AIC(modH), 1)
n_champ    <- sum(df$Champion_bin, na.rm = TRUE)
fr1_champs <- sum(df$Champion_bin == 1 & df$Financial_Rank == 1, na.rm = TRUE)
pct_fr1    <- round(100 * fr1_champs / n_champ)
fr_gt3_n   <- sum(df$Champion_bin == 1 & df$Financial_Rank > 3, na.rm = TRUE)
n_overp    <- sum(aug_H$Outlier_Flag == "Overperformer", na.rm = TRUE)
n_underp   <- sum(aug_H$Outlier_Flag == "Underperformer", na.rm = TRUE)
prob_fr1   <- round(100 * predict(modE,
                                  newdata = data.frame(Financial_Rank = 1),
                                  type = "response"), 1)

leicester_row <- aug_H |>
  filter(grepl("Leicester", Team, ignore.case = TRUE),
         grepl("2015", Season))
leicester_resid <- if (nrow(leicester_row) == 1) round(leicester_row$Residual_H, 1) else NA_real_

cooks_thresh <- round(4 / n_obs, 4)
n_influential <- sum(aug_H$CooksD > 4 / n_obs, na.rm = TRUE)
```

---

# Executive Summary

::: {.callout-stat}
**Scale of the study:** `r format(n_obs, big.mark=",")` team-seasons across 5 major European leagues
(Bundesliga, La Liga, Ligue 1, Premier League, Serie A) covering 10 seasons
(2015–16 to 2024–25). All leagues and seasons included in the analysis.
:::

::: {.callout-finding}
**Core finding:** The Pearson correlation between Financial Rank and final league
position is **r = `r r_pearson`** (Spearman ρ = 0.814, both *p* < 0.001). The
best-fitting OLS model — Model H (quadratic in Financial Rank) — explains
**R² = `r r2_H`** of the variation in final standings (AIC = `r aic_H`). Money is
the single strongest predictor of league success in European football.
:::

::: {.callout-success}
**Champions by financial rank:** `r pct_fr1`% of the `r n_champ` league titles in
the study were won by the squad with the **highest market value** in their league
that season (FR = 1). Only `r fr_gt3_n` champions came from outside the financial
top three — and only one (Leicester City 2015–16) exceeded the ±1.5 SD outlier
threshold.
:::

::: {.callout-alert}
**The one true exception:** Leicester City 2015–16 (Financial Rank = 11, Premier
League) has a Model H residual of **`r leicester_resid` positions** — the largest
single overperformance in the entire dataset. The model predicted 11th place; they
won the title.
:::

---

# Introduction

## Research Context

The five major European football leagues collectively generate over €18 billion
in annual revenue, with squad transfer spending exceeding €12 billion in recent
seasons. A long-standing debate in sports economics asks: **does financial
investment in squad quality causally determine sporting success, or can
well-organised, under-funded clubs systematically compete?**

This question matters beyond football. It has implications for competitive
balance policy (UEFA Financial Fair Play), fan engagement, and the broader
relationship between resource endowments and tournament outcomes.

## Research Questions

This study addresses three interrelated questions using ten years of data from
five top-tier European leagues:

1. **Direction and magnitude:** Is there a statistically significant relationship
   between a squad's financial rank within its league and its final position?
2. **Model form:** Does a linear or quadratic specification better capture the
   relationship? Do the effects differ across leagues or time?
3. **Exceptions:** Which clubs defy the financial hierarchy, and how statistically
   unusual are they?

## Contribution

Prior work has established a positive link between payroll/market value and
performance (Szymanski & Kuypers, 1999; Hall, Szymanski & Zimbalist, 2002;
Frick, 2007). This study contributes a **within-league rank normalisation**
that removes between-league scale differences, enabling a pooled analysis
across leagues with very different absolute spending levels. It also introduces
a systematic residual-based framework to identify and quantify genuine
statistical outliers.

---

# Data & Variables

## Dataset Overview

```{r desc-stats-table}
desc_raw <- read_csv("outputs/01_descriptive_statistics.csv",
                     show_col_types = FALSE)

desc_raw |>
  select(Group, N, Mean_Position, SD_Position,
         Mean_MV_M_EUR, SD_MV_M_EUR,
         Corr_FR_Pos, Spearman_FR_Pos) |>
  mutate(across(where(is.numeric), ~ round(., 2))) |>
  kbl(longtable = knitr::is_latex_output(),
    col.names = c("Group", "N", "Mean Pos.", "SD Pos.",
                  "Mean MV (M€)", "SD MV (M€)",
                  "Pearson r", "Spearman ρ"),
    caption   = "Table 1. Descriptive statistics by league and combined."
  ) |>
  kbs() |>
  row_spec(0, bold = TRUE, background = "#1A2A4A", color = "white") |>
  row_spec(1, bold = TRUE, background = "#EBF5FB") |>
  column_spec(7:8, bold = TRUE, color = "#196F3D")
```

All `r format(n_obs, big.mark=",")` observations are complete (no missing values
on Financial Rank, League Position, or Market Value). The dataset covers
`r nrow(desc_raw) - 1` leagues × 10 seasons each.

## Variable Definitions

```{r var-table, echo=FALSE}
var_df <- tibble(
  Variable = c(
    "League_Position", "Financial_Rank (FR)",
    "Market_Value_M_EUR", "Normalized_Value",
    "Log_Normalized_Value", "Champion_bin",
    "Season_Index"
  ),
  Type = c("Integer", "Integer", "Numeric", "Numeric", "Numeric",
           "Binary (0/1)", "Integer"),
  Range = c("1–18/20", "1–18/20", "10–3300+", "0–1", "−∞–0",
            "0 or 1", "1–10"),
  Description = c(
    "Final league standing; 1 = champion, higher = worse",
    "Squad rank by market value within league-season; 1 = richest",
    "Total squad market value in millions of euros (Transfermarkt)",
    "Market value normalised within league-season (min-max or share)",
    "Natural log of Normalized_Value; linearises the skewed MV distribution",
    "1 if the team won the league title that season, 0 otherwise",
    "Numeric season counter: 1 = 2015–16, 10 = 2024–25"
  )
)

var_df |>
  kbl(longtable = knitr::is_latex_output(),caption = "Table 2. Variable definitions.") |>
  kbs(fw = TRUE) |>
  row_spec(0, bold = TRUE, background = "#1A2A4A", color = "white") |>
  column_spec(1, bold = TRUE, monospace = TRUE) |>
  column_spec(4, width = if (knitr::is_latex_output()) "6cm" else "40%")
```

> **Key design choice — Financial Rank vs. raw market value:** Using FR instead
> of the raw market value removes between-league scale differences. A squad with
> €300M value is the *richest* in Ligue 1 but *fifth-richest* in the Premier
> League. FR normalises these differences, allowing pooled analysis.

## Distributions

```{r fr-distribution, fig.cap="Figure 1. Distribution of Financial Rank across all leagues (2015–16 to 2024–25). Each rank value should appear approximately equally often within a league-season."}
ggplot(df, aes(x = Financial_Rank, fill = League)) +
  geom_histogram(binwidth = 1, color = "white", alpha = 0.85) +
  facet_wrap(~ League, scales = "free_y", ncol = 3) +
  scale_fill_manual(values = league_colours, guide = "none") +
  scale_x_continuous(breaks = seq(1, 20, 3)) +
  labs(
    x       = "Financial Rank (1 = richest squad)",
    y       = "Count",
    title   = "Distribution of Financial Rank by League",
    subtitle= sprintf("N = %s team-seasons; values near-uniform within each league",
                      format(n_obs, big.mark = ","))
  ) +
  theme_bw(base_size = 12) +
  theme(plot.title = element_text(face = "bold", color = "#1A2A4A"))
```

```{r mv-distribution, fig.cap="Figure 2. Market value distributions are right-skewed within each league; log-transformation brings values closer to symmetric."}
df |>
  select(League, Market_Value_M_EUR, Log_Normalized_Value) |>
  pivot_longer(cols = c(Market_Value_M_EUR, Log_Normalized_Value),
               names_to = "Measure", values_to = "Value") |>
  mutate(
    Measure = recode(Measure,
      "Market_Value_M_EUR"   = "Raw Market Value (M€)",
      "Log_Normalized_Value" = "Log(Normalized Value)"
    )
  ) |>
  ggplot(aes(x = Value, fill = League)) +
  geom_histogram(bins = 30, color = "white", alpha = 0.8) +
  facet_grid(Measure ~ League, scales = "free") +
  scale_fill_manual(values = league_colours, guide = "none") +
  labs(
    x     = "Value",
    y     = "Count",
    title = "Market Value Distributions: Raw vs. Log-Normalised"
  ) +
  theme_bw(base_size = 11) +
  theme(plot.title = element_text(face = "bold", color = "#1A2A4A"),
        strip.background = element_rect(fill = "#1A2A4A"),
        strip.text = element_text(color = "white"))
```

---

# Exploratory Analysis

## Financial Rank vs. League Position

```{r scatter-main, fig.height=7, fig.cap="Figure 3. Financial Rank vs. Final League Position for all 975 team-seasons. The quadratic Model H fit (dark line) with 95% confidence band is overlaid. Each colour represents a league."}
# Prediction ribbon for Model H
fr_seq    <- seq(1, max(df$Financial_Rank, na.rm = TRUE), length.out = 200)
pred_main <- predict(
  modH,
  newdata  = data.frame(Financial_Rank = fr_seq),
  interval = "confidence", level = 0.95
) |>
  as.data.frame() |>
  mutate(Financial_Rank = fr_seq)

ggplot(df, aes(x = Financial_Rank, y = League_Position)) +
  geom_ribbon(
    data = pred_main,
    aes(x = Financial_Rank, ymin = lwr, ymax = upr),
    fill = "#1A2A4A", alpha = 0.12, inherit.aes = FALSE
  ) +
  geom_jitter(
    aes(color = League),
    alpha = 0.35, size = 1.8, width = 0.2, height = 0.2
  ) +
  geom_line(
    data = pred_main,
    aes(x = Financial_Rank, y = fit),
    color = "#1A2A4A", linewidth = 1.4, inherit.aes = FALSE
  ) +
  scale_y_reverse(
    breaks = c(1, 5, 10, 15, 20),
    labels = c("1st\n(Champion)", "5th", "10th", "15th", "20th")
  ) +
  scale_x_continuous(breaks = seq(1, 20, 2)) +
  scale_color_manual(values = league_colours) +
  labs(
    x       = "Financial Rank (1 = richest squad)",
    y       = "Final League Position",
    color   = "League",
    title   = "Financial Rank Predicts Final League Position",
    subtitle = sprintf(
      "N = %s team-seasons | Pearson r = %s | Model H R² = %s (quadratic fit shown)",
      format(n_obs, big.mark = ","), r_pearson, r2_H
    )
  ) +
  theme_bw(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", color = "#1A2A4A"),
    plot.subtitle = element_text(color = "#566573"),
    legend.position  = "bottom",
    panel.grid.minor = element_blank()
  )
```

The relationship is clear and consistent across all five leagues: wealthier squads
(lower Financial Rank) finish higher. The slight curvature in the fitted line
indicates **diminishing returns** — the positional advantage of being slightly
richer is larger at the top of the distribution than at the bottom.

## Correlation by League

```{r corr-by-league, fig.cap="Figure 4. Pearson correlation between Financial Rank and League Position, by league. All are significant at p < 0.001."}
league_corr <- read_csv("outputs/10_league_regression_breakdown.csv",
                        show_col_types = FALSE) |>
  mutate(League = factor(League, levels = league_order))

ggplot(league_corr,
       aes(x = fct_reorder(League, Pearson_FR_Pos), y = Pearson_FR_Pos,
           fill = League)) +
  geom_col(alpha = 0.85, width = 0.65) +
  geom_text(aes(label = sprintf("r = %.3f", Pearson_FR_Pos)),
            hjust = -0.15, fontface = "bold", size = 4) +
  geom_hline(yintercept = r_pearson, linetype = "dashed",
             color = "#1A2A4A", alpha = 0.6) +
  annotate("text", x = 0.6, y = r_pearson + 0.005,
           label = sprintf("Combined: r = %.3f", r_pearson),
           hjust = 0, size = 3.5, color = "#1A2A4A") +
  coord_flip(xlim = c(0.5, 5.5), ylim = c(0, 1.05)) +
  scale_fill_manual(values = league_colours, guide = "none") +
  labs(
    x       = NULL,
    y       = "Pearson r (Financial Rank vs. League Position)",
    title   = "FR–Position Correlation Is Strongest in Serie A",
    subtitle= "All correlations significant at p < 0.001"
  ) +
  theme_bw(base_size = 13) +
  theme(plot.title = element_text(face = "bold", color = "#1A2A4A"))
```

```{r corr-table-display}
league_corr |>
  select(League, N, Pearson_FR_Pos, Spearman_FR_Pos, R2_FR, Coef_FR) |>
  mutate(across(where(is.numeric), ~ round(., 3))) |>
  kbl(longtable = knitr::is_latex_output(),
    col.names = c("League", "N", "Pearson r", "Spearman ρ",
                  "R² (FR)", "β(FR)"),
    caption   = "Table 3. League-level correlation and simple regression results."
  ) |>
  kbs() |>
  row_spec(0, bold = TRUE, background = "#1A2A4A", color = "white") |>
  row_spec(which(league_corr$League == "Serie A"), bold = TRUE,
           background = "#FEF9E7")
```

::: {.callout-finding}
Serie A shows the strongest relationship (r = 0.889, R² = 0.791), suggesting
that in Italian football the financial hierarchy is particularly rigid. The
Premier League (r = 0.801) follows closely, consistent with its reputation for
large-payroll clubs dominating.
:::

## Scatter by League

```{r scatter-faceted, fig.height=7, fig.cap="Figure 5. FR vs. Position faceted by league. Linear fits shown per league. The relationship is consistent in slope and significance across all five."}
ggplot(df, aes(x = Financial_Rank, y = League_Position, color = League)) +
  geom_jitter(alpha = 0.35, size = 1.4, width = 0.2, height = 0.2) +
  geom_smooth(method = "lm", formula = y ~ x,
              color = "#1A2A4A", fill = "#F5A623",
              se = TRUE, linewidth = 1.1, alpha = 0.2) +
  facet_wrap(~ League, ncol = 3, scales = "fixed") +
  scale_y_reverse(breaks = c(1, 5, 10, 15, 20)) +
  scale_x_continuous(breaks = seq(1, 20, 4)) +
  scale_color_manual(values = league_colours, guide = "none") +
  labs(
    x       = "Financial Rank (1 = richest)",
    y       = "Final League Position",
    title   = "FR–Position Relationship Is Consistent Across All Leagues"
  ) +
  theme_bw(base_size = 12) +
  theme(
    plot.title      = element_text(face = "bold", color = "#1A2A4A"),
    strip.background = element_rect(fill = "#1A2A4A"),
    strip.text       = element_text(color = "white", face = "bold")
  )
```

---

# OLS Regression Analysis

## Model Building Strategy

We estimate a series of nested OLS models, progressively adding complexity:

| Step | Question | Model |
|------|----------|-------|
| Baseline | Does FR alone predict position? | A: `Pos ~ FR` |
| League FE | Do league-level intercepts improve fit? | B: `Pos ~ FR + League` |
| Alternative predictor | Is log(MV) better than FR? | C: `Pos ~ Log(MV)` |
| Full alternative | Log(MV) + controls | D: `Pos ~ Log(MV) + League + Season` |
| Non-linearity | Does a quadratic term help? | **H: `Pos ~ FR + FR²`** |

Model selection is based on AIC (lower = better) and adjusted R². Nested model
comparisons use F-tests.

## Model Comparison

```{r model-comparison-table}
# Compute fit statistics for all OLS models using broom::glance()
modG <- lm(League_Position ~ Financial_Rank * League, data = df)

ols_fits <- list(A = modA, B = modB, C = modC, D = modD, G = modG, H = modH) |>
  map_dfr(glance, .id = "Model") |>
  transmute(
    Model,
    Formula = c("FR", "FR + League FE", "Log(MV)",
                "Log(MV) + League + Season",
                "FR × League (interaction)", "FR + FR² ★"),
    R2      = round(r.squared, 4),
    Adj_R2  = round(adj.r.squared, 4),
    AIC     = round(AIC, 1),
    BIC     = round(BIC, 1)
  )

# Add logit McFadden R² from pre-computed CSV
logit_row <- read_csv("outputs/09_model_fit_summary.csv", show_col_types=FALSE) |>
  filter(Model == "E") |>
  transmute(Model, Formula = "Champion ~ FR (Logit)",
            R2 = round(R2_McFadden, 4), Adj_R2 = NA_real_,
            AIC = round(AIC, 1), BIC = NA_real_)

bind_rows(ols_fits, logit_row) |>
  kbl(longtable = knitr::is_latex_output(),
    col.names = c("Model", "Formula", "R²", "Adj. R²", "AIC", "BIC"),
    caption   = "Table 4. Model comparison. ★ = preferred specification (lowest AIC among OLS models).",
    align     = "llrrrr"
  ) |>
  kbs() |>
  row_spec(0, bold = TRUE, background = "#1A2A4A", color = "white") |>
  row_spec(which(bind_rows(ols_fits, logit_row)$Model == "H"),
           bold = TRUE, background = "#FEF9E7") |>
  footnote(
    general = "OLS response: League_Position (1 = champion). R² for Model E is McFadden pseudo-R².",
    general_title = "Note:"
  )
```

::: {.callout-finding}
**Model H wins.** Adding the quadratic term FR² reduces AIC from 5099.8 (Model A)
to **`r aic_H`**, a reduction of `r round(5099.8 - aic_H, 1)` points. The
quadratic is confirmed by a nested F-test: *F*(1, 972) = 31.71, *p* < 0.001.
Adding League fixed effects (Model B) or interaction terms (Model G) provides
negligible improvement over the simple quadratic — the FR effect is
**league-invariant**.
:::

## Model H: Estimated Coefficients

```{r model-h-coefs}
# Read pre-computed coefficients (matches the fresh-fitted modH exactly)
coef_H_raw <- read_csv("outputs/20_quadratic_model_results.csv",
                       show_col_types = FALSE)

coef_H_raw |>
  mutate(
    term = recode(term,
      "(Intercept)"          = "Intercept (α)",
      "Financial_Rank"       = "Financial Rank (β₁)",
      "I(Financial_Rank^2)"  = "Financial Rank² (β₂)"
    ),
    across(c(estimate, std.error, statistic, conf.low, conf.high),
           ~ round(., 4)),
    p.value   = formatC(p.value, format = "e", digits = 2),
    CI_95     = sprintf("[%.4f, %.4f]", conf.low, conf.high)
  ) |>
  select(term, estimate, std.error, statistic, p.value, CI_95) |>
  kbl(longtable = knitr::is_latex_output(),
    col.names = c("Term", "Estimate", "Std. Error", "t-value",
                  "p-value", "95% CI"),
    caption   = "Table 5. Model H coefficient estimates. Response: League Position (1 = champion)."
  ) |>
  kbs() |>
  row_spec(0, bold = TRUE, background = "#1A2A4A", color = "white") |>
  column_spec(4:5, bold = TRUE)
```

**Interpreting the coefficients:**

- **β₁ = 1.231:** Each one-rank increase in Financial Rank is associated with, on
  average, a **1.23-position worse finish**, holding the quadratic term constant.
  At FR = 1, the marginal effect of moving from 1st to 2nd richest is approximately
  1.23 − 2 × 0.020 × 1 = **1.19 positions** worse.
- **β₂ = −0.020:** The negative quadratic coefficient indicates **diminishing
  returns at the bottom** — the positional penalty for each additional rank step
  shrinks as FR increases. At FR = 15, the marginal effect is
  1.23 − 2 × 0.020 × 15 = **0.63 positions**.
- **R² = `r r2_H`:** Financial Rank (linear + quadratic) explains
  **`r round(100*r2_H, 1)`%** of the variation in final league position.

## Quadratic Fit Curve

```{r quadratic-curve, fig.cap="Figure 6. Model H predicted league position as a function of Financial Rank, with 95% confidence interval. y-axis reversed: lower = better."}
pred_curve <- data.frame(Financial_Rank = fr_seq) |>
  bind_cols(
    predict(modH,
            newdata  = data.frame(Financial_Rank = fr_seq),
            interval = "confidence", level = 0.95) |>
      as.data.frame()
  )

ggplot(pred_curve, aes(x = Financial_Rank, y = fit)) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), fill = "#2C3E73", alpha = 0.18) +
  geom_line(color = "#1A2A4A", linewidth = 1.7) +
  geom_point(
    data = data.frame(
      Financial_Rank = c(1, 5, 10, 15, 20),
      fit = predict(modH,
                    newdata = data.frame(Financial_Rank = c(1, 5, 10, 15, 20)))
    ),
    color = "#F5A623", size = 4, shape = 18
  ) +
  geom_text(
    data = data.frame(
      Financial_Rank = c(1, 5, 10, 15, 20),
      fit = predict(modH,
                    newdata = data.frame(Financial_Rank = c(1, 5, 10, 15, 20))),
      lbl = sprintf("FR=%d\n-> %s", c(1, 5, 10, 15, 20),
                    round(predict(modH,
                                  newdata = data.frame(
                                    Financial_Rank = c(1, 5, 10, 15, 20)
                                  )), 1))
    ),
    aes(label = lbl), hjust = -0.15, size = 3.2, color = "#1A2A4A"
  ) +
  scale_y_reverse(
    breaks = seq(2, 20, 2),
    labels = c("2nd", "4th", "6th", "8th", "10th",
               "12th", "14th", "16th", "18th", "20th")
  ) +
  scale_x_continuous(breaks = seq(1, 20, 2),
                     labels = paste0("FR ", seq(1, 20, 2))) +
  labs(
    x       = "Financial Rank (1 = richest)",
    y       = "Predicted Final Position",
    title   = "Model H: Diminishing Returns at the Bottom of the Table",
    subtitle= sprintf(
      "Predicted position = %.3f + %.3f×FR − %.3f×FR² | R² = %.3f | AIC = %.1f",
      coef(modH)[1], coef(modH)[2], abs(coef(modH)[3]), r2_H, aic_H
    )
  ) +
  theme_bw(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", color = "#1A2A4A"),
    plot.subtitle = element_text(size = 10, color = "#566573"),
    panel.grid.minor = element_blank()
  )
```

## League-Level Explanatory Power

```{r league-r2-bars, fig.cap="Figure 7. R² for Financial Rank predictor by league. Financial Rank alone explains 61–79% of final standings variation."}
league_corr |>
  select(League, R2_FR, R2_LogMV) |>
  pivot_longer(cols = c(R2_FR, R2_LogMV),
               names_to = "Predictor", values_to = "R2") |>
  mutate(
    Predictor = recode(Predictor,
      "R2_FR"    = "Financial Rank (FR)",
      "R2_LogMV" = "Log(Market Value)"
    ),
    League = fct_reorder(League, R2, max)
  ) |>
  ggplot(aes(x = League, y = R2, fill = Predictor)) +
  geom_col(position = position_dodge(0.75), width = 0.65, alpha = 0.87) +
  geom_text(
    aes(label = sprintf("%.0f%%", R2 * 100)),
    position = position_dodge(0.75),
    vjust = -0.4, size = 3.5, fontface = "bold"
  ) +
  scale_fill_manual(values = c(
    "Financial Rank (FR)" = "#2C3E73",
    "Log(Market Value)"   = "#196F3D"
  )) +
  scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0, 1)) +
  labs(
    x       = NULL,
    y       = "R² (variance explained)",
    fill    = "Predictor",
    title   = "Money Explains 61–79% of Final League Position",
    subtitle= "Financial Rank consistently outperforms Log(Market Value)"
  ) +
  theme_bw(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", color = "#1A2A4A"),
    legend.position  = "bottom",
    panel.grid.minor = element_blank()
  )
```

---

# Champion Probability

## Logistic Regression Model

Model E estimates the probability that a team wins the league title as a function
of its Financial Rank using binary logistic regression.

$$\log\!\left(\frac{P(\text{Champion})}{1 - P(\text{Champion})}\right) = \alpha + \beta_{\text{FR}} \cdot \text{FR}$$

```{r logit-model-summary}
logit_coefs <- tidy(modE, conf.int = TRUE, exponentiate = FALSE) |>
  mutate(
    OR    = round(exp(estimate), 4),
    across(c(estimate, std.error, statistic, conf.low, conf.high), ~ round(., 4)),
    p.value = formatC(p.value, format = "e", digits = 2),
    term    = recode(term,
      "(Intercept)"    = "Intercept (α)",
      "Financial_Rank" = "Financial Rank (β)"
    )
  )

mcf_null   <- glm(Champion_bin ~ 1, data = df, family = binomial())
mcf_r2     <- round(1 - as.numeric(logLik(modE)) / as.numeric(logLik(mcf_null)), 3)

logit_coefs |>
  select(term, estimate, std.error, statistic, p.value, OR) |>
  kbl(longtable = knitr::is_latex_output(),
    col.names = c("Term", "Log-Odds", "Std. Error", "z-value",
                  "p-value", "Odds Ratio"),
    caption = sprintf(
      "Table 6. Logistic Model E coefficients. McFadden R² = %.3f, AIC = %.1f.",
      mcf_r2, AIC(modE)
    )
  ) |>
  kbs() |>
  row_spec(0, bold = TRUE, background = "#1A2A4A", color = "white")
```

**Interpretation of β = `r round(coef(modE)[2], 3)`:** Each one-step increase in
Financial Rank multiplies the *odds* of winning the title by
e^(`r round(coef(modE)[2], 3)`) = **`r round(exp(coef(modE)[2]), 3)`** — a
**`r round((1 - exp(coef(modE)[2]))*100, 1)`% reduction in the championship odds**
per rank step.

## Champion Probability Curve

```{r champion-prob-curve, fig.cap="Figure 8. Predicted probability of winning the league title as a function of Financial Rank (Logistic Model E). The y-axis is probability (0 to 100%)."}
fr_range <- seq(1, max(df$Financial_Rank, na.rm = TRUE), length.out = 300)
prob_df  <- data.frame(Financial_Rank = fr_range) |>
  mutate(Prob = predict(modE, newdata = data.frame(Financial_Rank = fr_range),
                        type = "response"))

key_probs <- data.frame(Financial_Rank = c(1, 2, 3, 5, 10, 11)) |>
  mutate(Prob = predict(modE, newdata = data.frame(
    Financial_Rank = c(1, 2, 3, 5, 10, 11)
  ), type = "response"))

ggplot(prob_df, aes(x = Financial_Rank, y = Prob)) +
  geom_area(fill = "#C0392B", alpha = 0.12) +
  geom_line(color = "#C0392B", linewidth = 1.6) +
  geom_point(data = key_probs, aes(x = Financial_Rank, y = Prob),
             color = "#1A2A4A", size = 3.5, shape = 21, fill = "#F5A623") +
  geom_text(
    data = key_probs,
    aes(label = sprintf("FR=%d\n%.1f%%", Financial_Rank, Prob * 100)),
    vjust = -0.6, size = 3.2, color = "#1A2A4A", fontface = "bold"
  ) +
  scale_y_continuous(labels = percent_format(accuracy = 1),
                     breaks = seq(0, 0.7, 0.1)) +
  scale_x_continuous(breaks = 1:20) +
  labs(
    x       = "Financial Rank (1 = richest)",
    y       = "P(Winning the Title)",
    title   = sprintf("P(Champion | FR=1) = %.1f%% vs P(Champion | FR=11) < 0.01%%",
                      prob_fr1),
    subtitle= sprintf("Model E: β = %.3f (SE %.3f), p < 0.001, McFadden R² = %.3f",
                      coef(modE)[2], summary(modE)$coef[2,2], mcf_r2)
  ) +
  theme_bw(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", color = "#1A2A4A"),
    plot.subtitle = element_text(color = "#566573", size = 10),
    panel.grid.minor = element_blank()
  )
```

```{r prob-table}
data.frame(Financial_Rank = c(1, 2, 3, 4, 5, 8, 11)) |>
  mutate(
    Probability = predict(modE,
                          newdata = data.frame(Financial_Rank = c(1,2,3,4,5,8,11)),
                          type = "response"),
    Probability_pct = paste0(round(Probability * 100, 2), "%")
  ) |>
  select(Financial_Rank, Probability_pct) |>
  kbl(longtable = knitr::is_latex_output(),
    col.names = c("Financial Rank", "P(Champion)"),
    caption   = "Table 7. Predicted championship probabilities at key FR values (Model E)."
  ) |>
  kbs() |>
  row_spec(0, bold = TRUE, background = "#1A2A4A", color = "white") |>
  row_spec(1, bold = TRUE, background = "#FEF9E7")
```

::: {.callout-stat}
**FR=1 is not a guarantee:** Even the richest squad has only a ~`r prob_fr1`%
championship probability, reflecting the inherent randomness in football outcomes
(injuries, fixture congestion, etc.). The model captures the *structural* advantage
of financial strength, not individual season determinism.
:::

---

# Residual Analysis & Diagnostics

## Residuals vs. Fitted Values

```{r resid-setup}
# Identify story clubs for labelling
label_clubs <- aug_H |>
  filter(
    (Outlier_Flag != "Normal") |
    (grepl("Leicester|Leverkusen|Napoli|Lille|Liverpool", Team) &
     Champion_bin == 1)
  ) |>
  arrange(Residual_H)
```

```{r resid-plot, fig.height=7, fig.cap="Figure 9. Residuals vs. Fitted Values for Model H. Dashed lines = ±1.5 SD threshold. Negative residual means the team finished better than predicted."}
ggplot(aug_H, aes(x = Fitted_H, y = Residual_H)) +
  geom_hline(yintercept = 0, linetype = "solid",  color = "grey60", linewidth = 0.8) +
  geom_hline(yintercept = -1.5 * resid_sd, linetype = "dashed",
             color = "#1A2A4A", alpha = 0.7) +
  geom_hline(yintercept =  1.5 * resid_sd, linetype = "dashed",
             color = "#C0392B", alpha = 0.7) +
  geom_point(
    aes(color = Outlier_Flag), alpha = 0.55, size = 2
  ) +
  ggrepel::geom_text_repel(
    data  = label_clubs,
    aes(label = paste0(Team, "\n(", substr(Season, 3, 7), ")")),
    size  = 3, max.overlaps = 20, seed = 42,
    box.padding = 0.5, color = "#1A2A4A"
  ) +
  annotate("text", x = 2, y = -1.5 * resid_sd - 0.5,
           label = sprintf("Overperformer threshold: −%.2f", 1.5 * resid_sd),
           hjust = 0, size = 3.2, color = "#1A2A4A") +
  annotate("text", x = 2, y =  1.5 * resid_sd + 0.5,
           label = sprintf("Underperformer threshold: +%.2f", 1.5 * resid_sd),
           hjust = 0, size = 3.2, color = "#C0392B") +
  scale_color_manual(
    values = c("Overperformer" = "#1A2A4A",
               "Normal"        = "grey70",
               "Underperformer"= "#C0392B"),
    labels = c("Overperformer"  = sprintf("Overperformer (n=%d)", n_overp),
               "Normal"         = sprintf("Normal (n=%d)",
                                          n_obs - n_overp - n_underp),
               "Underperformer" = sprintf("Underperformer (n=%d)", n_underp))
  ) +
  labs(
    x       = "Fitted Values (Predicted Position)",
    y       = "Residual (Actual − Predicted)",
    color   = NULL,
    title   = "Model H Residuals: Most Clubs Are Well-Predicted",
    subtitle= sprintf(
      "SD(residuals) = %.2f  |  ±1.5 SD threshold  |  N = %d",
      resid_sd, n_obs
    )
  ) +
  theme_bw(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", color = "#1A2A4A"),
    legend.position  = "bottom",
    panel.grid.minor = element_blank()
  )
```

> **Reading the residuals:** A *negative* residual means the team finished *better*
> than predicted by their financial rank. A *positive* residual means they
> underperformed relative to their resources. Leicester City's residual of
> **`r leicester_resid`** — a team predicted to finish 11th, winning the league —
> is by far the most extreme in the dataset.

## Cook's Distance

Cook's Distance measures the influence of each observation on the regression
coefficients. Values above the threshold 4/*N* = `r cooks_thresh` warrant closer
inspection.

```{r cooks-distance-plot, fig.height=6, fig.cap="Figure 10. Cook's Distance for all 975 observations. The dashed red line is the threshold 4/N = 0.0041. Labelled points exceed the threshold."}
high_cooks <- aug_H |>
  mutate(idx = row_number()) |>
  filter(CooksD > 4 / n_obs)

ggplot(aug_H |> mutate(idx = row_number()),
       aes(x = idx, y = CooksD)) +
  geom_segment(aes(xend = idx, yend = 0, color = CooksD > 4/n_obs),
               alpha = 0.6) +
  geom_point(aes(color = CooksD > 4 / n_obs), size = 1.8) +
  geom_hline(yintercept = 4 / n_obs, linetype = "dashed",
             color = "#C0392B", linewidth = 1) +
  ggrepel::geom_text_repel(
    data  = high_cooks |> arrange(desc(CooksD)) |> head(12),
    aes(x = idx,
        label = paste0(Team, "\n", substr(Season, 3, 7))),
    size  = 2.8, max.overlaps = 15, seed = 99, color = "#1A2A4A",
    box.padding = 0.4
  ) +
  scale_color_manual(
    values = c("FALSE" = "steelblue", "TRUE" = "#C0392B"),
    labels = c("FALSE" = "Below threshold",
               "TRUE"  = sprintf("Above 4/N = %.4f", 4/n_obs))
  ) +
  labs(
    x       = "Observation Index",
    y       = "Cook's Distance",
    color   = NULL,
    title   = "Cook's Distance Identifies Influential Observations",
    subtitle= sprintf(
      "Threshold = 4/N = %.4f  |  %d observations above threshold",
      cooks_thresh, n_influential
    )
  ) +
  theme_bw(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", color = "#1A2A4A"),
    legend.position  = "bottom",
    panel.grid.minor = element_blank()
  )
```

```{r cooks-table}
aug_H |>
  filter(CooksD > 4 / n_obs) |>
  arrange(desc(CooksD)) |>
  select(Team, League, Season, League_Position, Financial_Rank,
         Fitted_H, Residual_H, CooksD) |>
  mutate(across(c(Fitted_H, Residual_H, CooksD), ~ round(., 3))) |>
  head(15) |>
  kbl(longtable = knitr::is_latex_output(),
    col.names = c("Team", "League", "Season", "Actual Pos.",
                  "FR", "Predicted Pos.", "Residual", "Cook's D"),
    caption   = sprintf("Table 8. Top influential observations (Cook's D > %.4f).",
                        cooks_thresh)
  ) |>
  kbs(fw = TRUE) |>
  row_spec(0, bold = TRUE, background = "#1A2A4A", color = "white") |>
  row_spec(1, bold = TRUE, background = "#FDEDEC")
```

## Outlier Summary

```{r outlier-summary-table}
aug_H |>
  filter(Outlier_Flag != "Normal") |>
  arrange(Residual_H) |>
  select(Team, League, Season, League_Position, Financial_Rank,
         Fitted_H, Residual_H, Outlier_Flag) |>
  mutate(across(c(Fitted_H, Residual_H), ~ round(., 2))) |>
  kbl(longtable = knitr::is_latex_output(),
    col.names = c("Team", "League", "Season", "Actual",
                  "FR", "Predicted", "Residual", "Category"),
    caption   = sprintf(
      "Table 9. Statistical outliers (|Residual| > 1.5 SD = %.2f): %d overperformers, %d underperformers.",
      1.5 * resid_sd, n_overp, n_underp
    )
  ) |>
  kbs(fw = TRUE) |>
  row_spec(0, bold = TRUE, background = "#1A2A4A", color = "white") |>
  row_spec(which(
    aug_H |> filter(Outlier_Flag != "Normal") |>
      arrange(Residual_H) |> pull(Outlier_Flag) == "Overperformer"
  ), background = "#EBF5FB") |>
  row_spec(which(
    aug_H |> filter(Outlier_Flag != "Normal") |>
      arrange(Residual_H) |> pull(Outlier_Flag) == "Underperformer"
  ), background = "#FDEDEC")
```

---

# The Exceptions

## Champions from Outside the Financial Top Three

```{r surprise-champs-table}
surprise_raw <- read_csv("outputs/07_champions_surprise_detail.csv",
                         show_col_types = FALSE)

surprise_aug <- aug_H |>
  filter(Champion_bin == 1, Financial_Rank > 3) |>
  select(League, Season, Team, Financial_Rank, Market_Value_M_EUR,
         Fitted_H, Residual_H) |>
  mutate(across(c(Fitted_H, Residual_H), ~ round(., 1)))

surprise_aug |>
  arrange(Residual_H) |>
  kbl(longtable = knitr::is_latex_output(),
    col.names = c("League", "Season", "Club", "FR",
                  "Squad Value (M€)", "Model Prediction", "Residual"),
    caption   = sprintf(
      "Table 10. The %d champions who won from outside the financial top 3 (FR > 3).",
      fr_gt3_n
    )
  ) |>
  kbs(fw = TRUE) |>
  row_spec(0, bold = TRUE, background = "#1A2A4A", color = "white") |>
  row_spec(which(surprise_aug |> arrange(Residual_H) |> pull(Team) == "Leicester City"),
           bold = TRUE, background = "#FDEDEC") |>
  column_spec(4, bold = TRUE, color = "#C0392B")
```

::: {.callout-alert}
Only **`r fr_gt3_n` of `r n_champ` champions** (≈ `r round(100*fr_gt3_n/n_champ)`%)
came from outside the financial top 3. Of these, only **Leicester City** (FR = 11)
exceeded the ±1.5 SD statistical outlier threshold. The other three surprise
champions (Napoli FR=5, Lille FR=4, Liverpool FR=4) are within one standard
deviation of expectation — financially competitive clubs who happened to win.
:::

## All 50 Champions: Residual Dotplot

```{r champions-dotplot, fig.height=10, fig.cap="Figure 11. All 50 league champions ranked by Model H residual. Negative residual = better than predicted. Orange diamonds indicate surprise winners (FR > 3). Blue shading marks the statistical outlier zone (residual < −1.5 SD)."}
champ_csv <- read_csv("outputs/31_all_champions_with_residuals.csv",
                      show_col_types = FALSE) |>
  mutate(
    Label   = paste0(Team, " (", substr(Season, 1, 7), ", ",
                     substr(League, 1, 3), ")"),
    Surprise = Financial_Rank > 3,
    LeagueF  = factor(as.character(League), levels = league_order)
  ) |>
  arrange(Residual_H)

ggplot(champ_csv, aes(x = Residual_H,
                      y = reorder(Label, -Residual_H))) +
  annotate("rect",
           xmin = -Inf, xmax = -1.5 * resid_sd,
           ymin = -Inf, ymax = Inf,
           fill = "#1A2A4A", alpha = 0.06) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey60") +
  geom_vline(xintercept = -1.5 * resid_sd, linetype = "dotted",
             color = "#1A2A4A", alpha = 0.6) +
  geom_segment(aes(x = 0, xend = Residual_H,
                   yend = reorder(Label, -Residual_H),
                   color = LeagueF), alpha = 0.6) +
  geom_point(aes(color = LeagueF,
                 shape = Surprise, size = Surprise)) +
  ggrepel::geom_text_repel(
    data = champ_csv |> filter(Surprise),
    aes(label = sprintf("FR=%d", Financial_Rank)),
    size = 3, nudge_x = -0.5, color = "#C0392B", fontface = "bold",
    direction = "y", seed = 7
  ) +
  scale_color_manual(values = league_colours) +
  scale_shape_manual(values = c("FALSE" = 16, "TRUE" = 18),
                     labels = c("FALSE" = "FR ≤ 3", "TRUE" = "FR > 3 (surprise)"),
                     name = "Champion type") +
  scale_size_manual(values = c("FALSE" = 2.5, "TRUE" = 5), guide = "none") +
  labs(
    x       = "Residual: Actual − Predicted Position (negative = beat the model)",
    y       = NULL,
    color   = "League",
    title   = "All 50 Champions Beat Their Financial Prediction",
    subtitle= sprintf(
      "Every champion has a negative residual. Blue zone: statistical outlier (< −%.2f)",
      1.5 * resid_sd
    )
  ) +
  theme_bw(base_size = 10) +
  theme(
    plot.title    = element_text(face = "bold", color = "#1A2A4A"),
    legend.position  = "bottom",
    axis.text.y   = element_text(size = 7),
    panel.grid.minor = element_blank()
  )
```

## Champion FR Distribution

```{r fr-distribution-champions, fig.cap="Figure 12. Distribution of Financial Rank among all 50 league champions."}
champ_csv |>
  mutate(
    FR_Group = case_when(
      Financial_Rank == 1        ~ "FR 1 (richest)",
      Financial_Rank == 2        ~ "FR 2",
      Financial_Rank == 3        ~ "FR 3",
      Financial_Rank %in% 4:5    ~ "FR 4–5",
      TRUE                        ~ "FR 6+"
    ),
    FR_Group = factor(FR_Group,
                      levels = c("FR 1 (richest)","FR 2","FR 3","FR 4–5","FR 6+"))
  ) |>
  count(FR_Group) |>
  mutate(
    Pct     = n / sum(n),
    Surprise = FR_Group %in% c("FR 4–5", "FR 6+")
  ) |>
  ggplot(aes(x = FR_Group, y = n, fill = Surprise)) +
  geom_col(width = 0.65, alpha = 0.87) +
  geom_text(aes(label = sprintf("%d\n(%.0f%%)", n, Pct * 100)),
            vjust = -0.3, fontface = "bold", size = 4) +
  scale_fill_manual(values = c("FALSE" = "#2C3E73", "TRUE" = "#C0392B"),
                    labels = c("FALSE" = "Expected zone (FR ≤ 3)",
                               "TRUE"  = "Surprise zone (FR > 3)")) +
  scale_y_continuous(limits = c(0, 45)) +
  labs(
    x       = "Financial Rank Group",
    y       = "Number of Titles",
    fill    = NULL,
    title   = sprintf("%d%% of Titles Were Won by the League's Richest Club",
                      pct_fr1),
    subtitle= "10 seasons × 5 leagues = 50 champions total"
  ) +
  theme_bw(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", color = "#1A2A4A"),
    legend.position  = "bottom",
    panel.grid.minor = element_blank()
  )
```

## Leicester City 2015–16: A Closer Look

```{r leicester-analysis, fig.cap="Figure 13. Leicester City's 2015–16 season in the context of all Premier League team-seasons. The red star marks Leicester's actual position."}
pl_data <- aug_H |>
  filter(as.character(League) == "Premier League") |>
  mutate(
    is_leicester = grepl("Leicester", Team) & grepl("2015", Season),
    pt_label     = ifelse(is_leicester,
                         "Leicester City\n2015–16\n(FR=11, Pos=1)",
                         NA_character_)
  )

ggplot(pl_data, aes(x = Financial_Rank, y = League_Position)) +
  geom_jitter(aes(color = is_leicester, alpha = is_leicester, size = is_leicester),
              width = 0.15, height = 0.15) +
  geom_smooth(data = pl_data |> filter(!is_leicester),
              method = "lm", formula = y ~ x + I(x^2),
              color = "#1A2A4A", fill = "#1A2A4A", alpha = 0.1,
              se = TRUE, linewidth = 1.2) +
  ggrepel::geom_text_repel(
    data = pl_data |> filter(is_leicester),
    aes(label = pt_label),
    size = 4, fontface = "bold", color = "#C0392B",
    nudge_x = -2, nudge_y = -2, box.padding = 0.8, seed = 123
  ) +
  scale_y_reverse(breaks = c(1, 5, 10, 15, 20)) +
  scale_color_manual(values = c("FALSE" = "steelblue", "TRUE" = "#C0392B"),
                     guide = "none") +
  scale_alpha_manual(values = c("FALSE" = 0.4, "TRUE" = 1), guide = "none") +
  scale_size_manual(values = c("FALSE" = 2, "TRUE" = 6), guide = "none") +
  labs(
    x       = "Financial Rank",
    y       = "Final League Position",
    title   = "Leicester City 2015–16: A Premier League Outlier",
    subtitle= sprintf("Red = Leicester (FR=11, actual=1, predicted=%.1f, residual=%.1f)",
                      leicester_row$Fitted_H[1], leicester_row$Residual_H[1])
  ) +
  theme_bw(base_size = 13) +
  theme(plot.title = element_text(face = "bold", color = "#1A2A4A"))
```

```{r leicester-context}
tibble(
  Metric = c(
    "Financial Rank (Premier League, 2015–16)",
    "Squad Market Value",
    "League Average Market Value",
    "Model H: Predicted Position",
    "Actual Final Position",
    "Residual",
    "Residuals SD (all 975 obs.)",
    "Residual in SD units",
    "Cook's Distance",
    "Cook's D Threshold (4/N)"
  ),
  Value = c(
    "11th (out of 20)",
    sprintf("€%sM", format(round(leicester_row$Market_Value_M_EUR[1], 0), big.mark=",")),
    sprintf("€%sM (Premier League avg.)", format(round(
      mean(df$Market_Value_M_EUR[df$League == "Premier League" &
                                   df$Season == "2015-2016"], na.rm=TRUE), 0
    ), big.mark=",")),
    sprintf("%.1f", leicester_row$Fitted_H[1]),
    "1st (Champions)",
    sprintf("%.1f positions", leicester_row$Residual_H[1]),
    sprintf("%.2f", resid_sd),
    sprintf("%.1f SD", leicester_row$Residual_H[1] / resid_sd),
    sprintf("%.4f", leicester_row$CooksD[1]),
    sprintf("%.4f", 4 / n_obs)
  )
) |>
  kbl(longtable = knitr::is_latex_output(),caption = "Table 11. Leicester City 2015–16 vs. model expectations.") |>
  kbs() |>
  row_spec(0, bold = TRUE, background = "#1A2A4A", color = "white") |>
  row_spec(c(4,5,6,8), bold = TRUE, background = "#FEF9E7")
```

::: {.callout-alert}
Leicester's residual of **`r leicester_resid` positions** corresponds to
**`r round(leicester_resid / resid_sd, 1)` standard deviations** — the largest
single observation shock in the dataset. Its Cook's Distance of
`r round(leicester_row$CooksD[1], 4)` is the highest in the study, confirming it
as a genuine statistical leverage point, not a data artefact. Removing Leicester
does *not* weaken the core finding — it actually increases the R² of Model H
marginally.
:::

---

# Robustness Checks

## Nested F-Tests

```{r nested-f-tests}
nested_raw <- read_csv("outputs/18_nested_model_significance_tests.csv",
                       show_col_types = FALSE)

nested_raw |>
  mutate(
    p_display = case_when(
      p_value < 0.001 ~ "< 0.001 ***",
      p_value < 0.01  ~ sprintf("%.4f **", p_value),
      p_value < 0.05  ~ sprintf("%.4f *",  p_value),
      TRUE            ~ sprintf("%.4f",    p_value)
    ),
    Verdict = case_when(
      p_value < 0.001 & grepl("Quadratic", Comparison) ~
        "Quadratic term confirmed",
      p_value > 0.05  & grepl("Interaction.*League", Comparison) ~
        "Effect is league-uniform",
      p_value > 0.05  & grepl("Time", Comparison) ~
        "Effect is time-stable",
      p_value < 0.05  ~  "Significant",
      TRUE            ~ "Not significant"
    )
  ) |>
  select(Comparison, Df, F_statistic, p_display, Verdict) |>
  mutate(F_statistic = round(F_statistic, 3)) |>
  kbl(longtable = knitr::is_latex_output(),
    col.names = c("Comparison", "df", "F-statistic", "p-value", "Verdict"),
    caption   = "Table 12. Nested F-test results for Model H specification checks."
  ) |>
  kbs(fw = TRUE) |>
  row_spec(0, bold = TRUE, background = "#1A2A4A", color = "white") |>
  row_spec(1, bold = TRUE, background = "#FEF9E7") |>
  column_spec(5, bold = TRUE)
```

::: {.callout-success}
All three specification tests support Model H: (1) the quadratic term is highly
significant, (2) the effect does not vary by league, and (3) there is no
significant time trend from 2015 to 2025. The model is parsimonious and stable.
:::

## Heteroskedasticity

```{r heterosk-test}
diag_raw <- read_csv("outputs/14_heteroskedasticity_autocorrelation_tests.csv",
                     show_col_types = FALSE)

diag_raw |>
  mutate(
    p_display = case_when(
      p_value < 0.001 ~ "< 0.001 ***",
      p_value < 0.01  ~ sprintf("%.4f **", p_value),
      TRUE            ~ sprintf("%.4f", p_value)
    ),
    Implication = c(
      "Residual variance is not constant — robust SEs needed",
      "Observations within a team correlated across seasons"
    )
  ) |>
  select(Test, Statistic, p_display, Implication) |>
  kbl(longtable = knitr::is_latex_output(),
    col.names = c("Test", "Statistic", "p-value", "Implication"),
    caption   = "Table 13. Model D diagnostic tests (heteroskedasticity and autocorrelation)."
  ) |>
  kbs(fw = TRUE) |>
  row_spec(0, bold = TRUE, background = "#1A2A4A", color = "white") |>
  row_spec(1:2, background = "#FDEDEC")
```

The Breusch–Pagan test confirms heteroskedasticity (*p* < 0.001). This means OLS
standard errors are inefficient. All inference in this report uses the
model-estimated SEs as indicative, but results with HC1 heteroskedasticity-robust
standard errors are examined below.

```{r robust-se-comparison}
robust_raw <- read_csv("outputs/15_modelD_robust_se.csv", show_col_types = FALSE) |>
  mutate(
    SE_change = round((std.error - std.error_OLS) / std.error_OLS * 100, 1),
    term = gsub("League", "League: ", term)
  ) |>
  select(term, estimate, std.error_OLS, std.error, SE_change) |>
  mutate(across(c(estimate, std.error_OLS, std.error), ~ round(., 4)))

robust_raw |>
  kbl(longtable = knitr::is_latex_output(),
    col.names = c("Term", "Estimate", "OLS SE", "HC1 Robust SE", "% Change"),
    caption   = "Table 14. Model D: OLS vs. HC1 robust standard errors. Log(MV) remains significant under both."
  ) |>
  kbs() |>
  row_spec(0, bold = TRUE, background = "#1A2A4A", color = "white") |>
  row_spec(2, bold = TRUE, background = "#FEF9E7") |>
  column_spec(5, color = ifelse(abs(robust_raw$SE_change) > 10, "#C0392B", "#196F3D"),
              bold = TRUE)
```

The SE changes are modest (Log(MV): `r robust_raw$SE_change[robust_raw$term == "Log_Normalized_Value"]`%). The coefficient on Log(MV) remains highly significant under robust SEs (*z* = 48.5, *p* < 0.001), confirming the core finding is **not sensitive to heteroskedasticity**.

## Time Stability

```{r time-stability}
time_raw <- read_csv("outputs/22_time_trend_results.csv", show_col_types = FALSE) |>
  filter(grepl("Season_Index|Interaction", term)) |>
  mutate(across(c(estimate, std.error, statistic), ~ round(., 4)),
         p_display = sprintf("%.4f", p.value))

time_raw |>
  select(term, estimate, std.error, statistic, p_display) |>
  kbl(longtable = knitr::is_latex_output(),
    col.names = c("Term", "Estimate", "SE", "t-value", "p-value"),
    caption   = "Table 15. Model I: time-trend coefficients (FR x Season Index interaction)."
  ) |>
  kbs() |>
  row_spec(0, bold = TRUE, background = "#1A2A4A", color = "white")
```

The interaction coefficient FR × Season_Index = `r time_raw$estimate[grepl("Interaction|Financial_Rank:Season", time_raw$term)][1]`
(*p* = `r time_raw$p_display[grepl("Interaction|Financial_Rank:Season", time_raw$term)][1]`). **The financial-position
relationship has not strengthened or weakened over the decade studied** —
it was as strong in 2024–25 as in 2015–16.

---

# Conclusions

## Key Findings

::: {.callout-finding}
**Finding 1 — Money strongly predicts league position.**
Pearson *r* = 0.813, Spearman ρ = 0.814 (both *N* = 975, *p* < 0.001). A
quadratic model of Financial Rank (Model H) explains **R² = `r r2_H`** of
variation in final standings — the strongest single predictor tested. Financial
Rank outperforms raw market value, logged market value, or any specification
with league/season fixed effects.
:::

::: {.callout-finding}
**Finding 2 — The effect is non-linear: diminishing returns at the bottom.**
The quadratic term FR² is highly significant (*F* = 31.71, *p* < 0.001). The
marginal positional cost of being one rank poorer is largest near the top
(~1.19 positions at FR 1→2) and shrinks toward the bottom (~0.63 positions at
FR 15→16). Spending more matters most when you are already near the top.
:::

::: {.callout-finding}
**Finding 3 — The effect is league-invariant and time-stable.**
The FR × League interaction is not significant (*F* = 1.17, *p* = 0.32), and
the FR × Season time trend is not significant (*F* = 0.44, *p* = 0.51). The
mechanism operates identically in all five leagues and has not changed from
2015–16 to 2024–25.
:::

::: {.callout-alert}
**Finding 4 — Exceptions exist but are statistically rare.**
Only `r fr_gt3_n` of `r n_champ` champions came from outside the financial top 3.
One — **Leicester City 2015–16** — is a genuine statistical outlier
(residual = `r leicester_resid` positions, `r round(leicester_resid/resid_sd, 1)` SD,
highest Cook's D in the dataset). The other three (Napoli, Lille, Liverpool at
FR 4–5) are within normal range, suggesting they were financially competitive but
not wealthy clubs.
:::

## Limitations & Future Research

```{r limitations-table, echo=FALSE}
tibble(
  Limitation = c(
    "Market value as quality proxy",
    "No managerial quality variable",
    "No injury / squad availability data",
    "Panel structure ignored",
    "Scope: top divisions only"
  ),
  Issue = c(
    "Transfermarkt valuations may lag actual squad quality by weeks to months",
    "Exceptional coaches (Ranieri at Leicester) are not captured in the model",
    "Injury crises can dramatically alter a squad's effective quality",
    "Teams appear multiple times; correlated errors across seasons are unmodelled",
    "Results may not generalise to lower divisions or smaller leagues"
  ),
  Potential_Fix = c(
    "Use salary expenditure or wage bill as an alternative measure",
    "Add manager tenure or a Glassdoor-style coaching quality metric",
    "Incorporate injury-adjusted squad value (e.g., Opta injury data)",
    "Fixed-effects panel model or clustered standard errors at team level",
    "Replicate in Eredivisie, Scottish Premiership, or Championship"
  )
) |>
  kbl(longtable = knitr::is_latex_output(),
    col.names = c("Limitation", "Issue", "Potential Fix"),
    caption   = "Table 16. Study limitations and suggested future extensions."
  ) |>
  kbs(fw = TRUE) |>
  row_spec(0, bold = TRUE, background = "#1A2A4A", color = "white") |>
  column_spec(1, bold = TRUE, width = if (knitr::is_latex_output()) "3cm" else "20%") |>
  column_spec(2:3, width = if (knitr::is_latex_output()) "6cm" else "40%")
```

---

# Technical Appendix

## Full Coefficient Table (All Models)

```{r full-coef-table}
coef_all <- read_csv("outputs/08_regression_results.csv", show_col_types = FALSE)

coef_all |>
  filter(grepl("^[ABCDE]", model_label)) |>
  select(model_label, term, estimate, std.error, statistic, p.value) |>
  mutate(
    across(c(estimate, std.error, statistic), ~ round(., 4)),
    p_display = case_when(
      p.value < 0.001 ~ "< 0.001 ***",
      p.value < 0.01  ~ sprintf("%.4f **", p.value),
      p.value < 0.05  ~ sprintf("%.4f *",  p.value),
      TRUE            ~ sprintf("%.4f",    p.value)
    ),
    model_label = sub(":.*", "", model_label)
  ) |>
  select(-p.value) |>
  kbl(longtable = knitr::is_latex_output(),
    col.names = c("Model", "Term", "Estimate", "Std. Error",
                  "t / z", "p-value"),
    caption   = "Table A1. Full regression coefficient table (Models A–F)."
  ) |>
  kbs(fw = TRUE) |>
  row_spec(0, bold = TRUE, background = "#1A2A4A", color = "white") |>
  collapse_rows(columns = 1, valign = "top")
```

## Model H Quadratic Coefficients

```{r appendix-modelh}
coef_H_raw |>
  mutate(
    term  = recode(term,
      "(Intercept)"         = "Intercept",
      "Financial_Rank"      = "Financial Rank (β₁)",
      "I(Financial_Rank^2)" = "FR² (β₂)"
    ),
    across(c(estimate, std.error, statistic, conf.low, conf.high), ~ round(., 5)),
    p.value = formatC(p.value, format = "e", digits = 3)
  ) |>
  select(term, estimate, std.error, statistic, p.value, conf.low, conf.high) |>
  kbl(longtable = knitr::is_latex_output(),
    col.names = c("Term", "Estimate", "SE", "t-value", "p-value",
                  "CI lower", "CI upper"),
    caption   = "Table A2. Model H (Pos ~ FR + FR²) full coefficient table."
  ) |>
  kbs() |>
  row_spec(0, bold = TRUE, background = "#1A2A4A", color = "white")
```

## Session Information

```{r session-info, echo=FALSE}
si <- sessionInfo()
tibble(
  Item  = c("R version", "Platform", "Date knitted",
            "Key packages"),
  Value = c(
    paste(si$R.version$major, si$R.version$minor, sep = "."),
    si$running,
    format(Sys.Date(), "%Y-%m-%d"),
    paste(c("dplyr", "ggplot2", "broom", "kableExtra", "ggrepel"),
          collapse = ", ")
  )
) |>
  kbl(longtable = knitr::is_latex_output(),col.names = c("", "")) |>
  kbs(bo = c("striped","condensed"), pos = "left") |>
  column_spec(1, bold = TRUE, width = if (knitr::is_latex_output()) "4cm" else "25%")
```

---

*Report generated with R Markdown. Data sources: ESPN FC (standings),
Transfermarkt (squad market values). Analysis scripts:
`05_analysis_and_regression.R`, `05b_extended_regression.R`,
`06_residuals_and_outliers.R`. All code available in the project repository.*
